### 2022-03-22 omicron-generation.tex
spelling check

Estimating the generation-intervaldistribution is challenging, in part, due to difficulties in observing actual infection events.
Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
Each of these transmission-interval distributions are subject to dynamical effects which, can bias estimation.
For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
Therefore, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions, especially if their growth rates differ.

@@

While the derivation of the forward serial-interval distribution \eref{forwardserial} may be complex, its implementation is simple.
The main difference between our model and prevous models that neglect dynamical effects \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation} is the exponential growth term $\exp(r \alpha_1)$ and the normalization term $\phi$---it is relatively straightforward to include these terms in existing models of serial intervals.
\cite{ferretti2020quantifying, ferretti2020timing} also included this term in their analyses of serial-interval data, but only accounted for the epidemic growth effect (and not the decay effect).

@@

Similar pictures arise for between-household transmission pairs, but the differencse in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
Consistent with the original study, which also reported shorter mean serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean generation intervals for between-household Delta pairs.
While the difference in mean generation intervals is larger, there is greater uncertainty in their mean estimates (\fref{serial}F): 2.9 days (95\% CI: 2.5--3.3 days) for the Omicron variant and 3.5 days (95\% CI: 3.2--3.8 dyas) for the Delta variant.
Once again, these patterns are robust across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants (\fref{serial}G--H).

@@

A few studies have suggesed the the incubation period of the Omicron variant may be shorter than that of the Delta variant.
The median estimates of the Omicron incubation period typically range between 3--4 days, consistent with earlier findings of \citep{backer2021omicron}. 
However, these data were collected when the number of Omicron infections was growing rapidly \citep{jansen2021investigation,song2022serial}, suggesting that they may have been subject to similar biases.
On the other hand, incubation-period estimates based on individuals who were exposed from the same event are likely more reliable (because they look forward in time):
\cite{brandal2021outbreak} estimated the median incubation period of the Omicron variant to be 3 days among those who attended the same holiday party ($n=117$) on 26 November 2021 in Norway. 
However, we cannot rule out the possibility that some of these attendees were infected prior to the party given that some individuals had COVID-like symptoms prior to the party with at least 96 of the attendees sharing offices; neglecting these factors can lead to underestimation of the mean incubation period.
Systematic comparisons of data collection methods and epidemiological contexts are needed to properly assess the differences in incubation period distributions of the Delta and Omicron variants.

A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron varaint but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and faster within-host clearance of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of generation intervals.

@@

We considered two ways of measuring transmission advantanges: growth-rate differences and reproduction advantage.
Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
On the other hand, the growth-rate difference likely provides a better measure for transmission advantage for an ongoing epidemic as it provides information about which strain is growing faster.
For example, when two strains have the same $\RR$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
Therefore, we suggest using growth-rate differences and reproduction advantage as complementary measures for understanding the dynamics of emerging SARS-CoV-2 variants.

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_stratified.pdf}
\caption{
\textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-househol transmission) and the week of infectors' symptom onset (weeks 50 vs 51 of 2021).
Lines represent maximum likelihood estimates.
Shaded regions represent the corresponding 95\% confidence intervals.
}
\end{figure}

>>>>

Estimating the generation-interval distribution is challenging, in part, due to difficulties in observing actual infection events.
Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
Each of these transmission-interval distributions are subject to dynamical effects which, can bias estimation.
For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
Therefore, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions, especially if their growth rates differ.

@@

While the derivation of the forward serial-interval distribution \eref{forwardserial} may be complex, its implementation is simple.
The main difference between our model and previous models that neglect dynamical effects \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation} is the exponential growth term $\exp(r \alpha_1)$ and the normalization term $\phi$---it is relatively straightforward to include these terms in existing models of serial intervals.
\cite{ferretti2020quantifying, ferretti2020timing} also included this term in their analyses of serial-interval data, but only accounted for the epidemic growth effect (and not the decay effect).

@@

Similar pictures arise for between-household transmission pairs, but the differences in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
Consistent with the original study, which also reported shorter mean serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean generation intervals for between-household Delta pairs.
While the difference in mean generation intervals is larger, there is greater uncertainty in their mean estimates (\fref{serial}F): 2.9 days (95\% CI: 2.5--3.3 days) for the Omicron variant and 3.5 days (95\% CI: 3.2--3.8 days) for the Delta variant.
Once again, these patterns are robust across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants (\fref{serial}G--H).

@@

A few studies have suggested the the incubation period of the Omicron variant may be shorter than that of the Delta variant.
The median estimates of the Omicron incubation period typically range between 3--4 days, consistent with earlier findings of \citep{backer2021omicron}. 
However, these data were collected when the number of Omicron infections was growing rapidly \citep{jansen2021investigation,song2022serial}, suggesting that they may have been subject to similar biases.
On the other hand, incubation-period estimates based on individuals who were exposed from the same event are likely more reliable (because they look forward in time):
\cite{brandal2021outbreak} estimated the median incubation period of the Omicron variant to be 3 days among those who attended the same holiday party ($n=117$) on 26 November 2021 in Norway. 
However, we cannot rule out the possibility that some of these attendees were infected prior to the party given that some individuals had COVID-like symptoms prior to the party with at least 96 of the attendees sharing offices; neglecting these factors can lead to underestimation of the mean incubation period.
Systematic comparisons of data collection methods and epidemiological contexts are needed to properly assess the differences in incubation period distributions of the Delta and Omicron variants.

A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and faster within-host clearance of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of generation intervals.

@@

We considered two ways of measuring transmission advantages: growth-rate differences and reproduction advantage.
Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
On the other hand, the growth-rate difference likely provides a better measure for transmission advantage for an ongoing epidemic as it provides information about which strain is growing faster.
For example, when two strains have the same $\RR$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
Therefore, we suggest using growth-rate differences and reproduction advantage as complementary measures for understanding the dynamics of emerging SARS-CoV-2 variants.

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_stratified.pdf}
\caption{
\textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-household transmission) and the week of infectors' symptom onset (weeks 50 vs 51 of 2021).
Lines represent maximum likelihood estimates.
Shaded regions represent the corresponding 95\% confidence intervals.
}
\end{figure}
